# Exercise #1: Loading Spatial Data

## Load Shapefiles Into R

The `sf` package has wide-ranging support for importing various spatial
data/file formats. You can even import data directly from a spatially
enabled database (e.g. PostgreSQL/PostGIS). One of the more common spatial
file formats you may encounter is the ESRI Shapefile format.

Here, we will work with an example from a research effort surveying for
_Franciscana_. Prior to the survey effort, there was a region identified as
a hiatus, or gap, where there were no historical sightings of _Franciscana_. This
region was defined and is provided as a Shapefile.

```{r load-franciscana}
library(tidyverse)
library(sf)
library(lwgeom)
```

provide the path to your shapefile directory. Note the path is based
on the *.shp file.

```{r}
hiatus_path <- file.path("..","data","Fraciscana_Example","Hiatus.shp")

hiatus <- sf::st_read(hiatus_path)
```

This demonstrates a common issue with Shapefiles. Sometimes they can be improperly
formed or corrupted. The `lwgeom` package has a function `st_make_valid()` which
can fix most of these issues. There is one additional step required
(`sf::st_collection_extrac()`) which simplifies the data object from a 
"GEOMETRYCOLLECTION" to a single "POLYGON". Don't get too hung up on these steps
to 'fix' the Shapefile data. Unfortunately, it is just one of those things that
you might have to wrestle with when importing data. `st_make_valid()` and other
functions within R are helpful. But, you may still need to spend some time
Googling the error message or asking for help on Stack Overflow or Twitter.

Note that we are using the _pipe_ (`%>%`) function in the code below. This is a
key component of the _tidyverse_ and may be new to users more familiar with
standard R code. 

```{r, make-valid}
library(lwgeom)
hiatus <- hiatus %>% lwgeom::st_make_valid() %>% 
  sf::st_collection_extract()
# base R version
# hiatus <- lwgeom::st_make_valid(hiatus)
```

Now, let's just do a quick plot of hiatus so we can verify it looks valid

```{r, check-plot}
hiatus %>% plot()
```

## First Map with ggplot

While this demonstrates that we have successfully imported the spatial data into
R, it is not a very satisfying map. We will use the `ggplot2` package as the
basis for all of our mapping. If you are not already familiar with `ggplot2`, it
is a widely used framework for creating all types of plots and graphics from
within R. `ggplot2` has special extensions and capabilities with respect to 
simple features (`sf`) data that simplifies creation of publication quality
maps.

First, lets repeat our previous plot but this time with `ggplot2`

```{r quick_ggplot}
# ggplot2 is already loaded as part of the tidyverse
ggplot() + geom_sf(data = hiatus)
```

This is a nicer looking plot and, note, `ggplot2` has kindly added graticule
ticks for the longitude and latitude values. It would be nice, however, if we
had some land and ocean features so we can better appreciate this hiatus region.

For this, we can use the `ggspatial` package to import map tiles hosted exernally
via the internet. In this first example, we will pull map tiles from the Open
Street Map service.

```{r hiatus-osm}
library(ggspatial)

ggplot() +
  annotation_map_tile(type = "osm") +
  layer_spatial(hiatus)
```

This is an improvement, but it might be better if we added some transparency
to the hiatus polygon so we can see the land features underneath. We can also
change the color and weight of the polygon outline so it stands out more.

```{r hiatus-osm-2}
ggplot() +
  annotation_map_tile(type = "osm") +
  layer_spatial(hiatus, col = "dark red", alpha = 0.4, size = 2)
```

Now we have a slightly better sense of where we are in relation to the nearby
land. But, it might be useful to zoom out a bit more so we can get a better
sense of the areas location and nearby geograph. Note the arguments to 
_expand_ within `scale_x_continuous` and `scale_y_continuous` are in units
of degrees longitude and latitude.

While we are at it, we can also add a scale bar in the bottom right.

```{r hiatus-osm-3}
ggplot() +
  annotation_map_tile(type = "osm") +
  layer_spatial(hiatus, col = "dark red", alpha = 0.4, size = 1) +
  scale_y_continuous(expand = c(1.5,1.5)) +
  scale_x_continuous(expand = c(2,2)) +
  annotation_scale(location = "br", width_hint = 0.5, style = "ticks")
```

The Open Street Map project is not really designed for environmental or marine
centric maps. A better option might be the ESRI World Ocean Basemap which provides
detailed bathymetry and topography for the entire globe.

We need to specify the URL location for the tile service:

```{r esri-ocean-tiles}
esri_ocean <- paste0('https://services.arcgisonline.com/arcgis/rest/services/',
                     'Ocean/World_Ocean_Base/MapServer/tile/${z}/${y}/${x}.jpeg')
```

And, then we can replace the `annotation_map_tile()` in the previous code

```{r hiatus-esri-ocean}
ggplot() +
  annotation_map_tile(type = esri_ocean, zoomin = 1, progress = "none") +
  layer_spatial(hiatus, col = "dark red", alpha = 0.4, size = 1) +
  scale_y_continuous(expand = c(1.5,1.5)) +
  scale_x_continuous(expand = c(2,2)) +
  annotation_scale(location = "br", width_hint = 0.5, style = "ticks")
```

Now that we no longer have place names, it is important to add a title and
subtitle to the map so we can inform the user about what this is and where
it is in the world.

```{r hiatus-esri-ocean-2}
ggplot() +
  annotation_map_tile(type = esri_ocean, zoomin = 1, progress = "none") +
  layer_spatial(hiatus, col = "dark red", alpha = 0.4, size = 1) +
  scale_y_continuous(expand = c(1.5,1.5)) +
  scale_x_continuous(expand = c(2,2)) +
  annotation_scale(location = "tl", 
                   width_hint = 0.5, 
                   style = "ticks") +
  ggtitle("Franciscana Hiatus Region",
          subtitle = "Area of of NE Brazil with no Franciscana sightings")
```

There are variety of other style options available for tile maps that you
can explore. Here are some additional tiles that you might prefer

### cartolight

```{r hiatus-cartolight}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 1, progress = "none") +
  layer_spatial(hiatus, col = "dark red", alpha = 0.4, size = 1) +
  scale_y_continuous(expand = c(1.5,1.5)) +
  scale_x_continuous(expand = c(2,2)) +
  annotation_scale(location = "tl", 
                   width_hint = 0.5, 
                   style = "ticks") +
  ggtitle("Franciscana Hiatus Region",
          subtitle = "Area of of NE Brazil with no Franciscana sightings")
```

### Microsoft Bing Aerial Imagery

```{r hiatus-bing-aerial}
ggplot() +
  annotation_map_tile(type = "http://ecn.t3.tiles.virtualearth.net/tiles/a${q}.jpeg?g=1", zoomin = 1, progress = "none") +
  layer_spatial(hiatus, col = "dark red", alpha = 0.4, size = 1) +
  scale_y_continuous(expand = c(1.5,1.5)) +
  scale_x_continuous(expand = c(2,2)) +
  annotation_scale(location = "tl", 
                   width_hint = 0.5, 
                   style = "ticks") +
  ggtitle("Franciscana Hiatus Region",
          subtitle = "Area of of NE Brazil with no Franciscana sightings")
```

### Projections

The data we imported from the shapefile is in lat-long coordinates. This is a convenient
coordinate system and is the common system for storing geographic coordinates the world
over. GPS, Satellite Telemetry, and many other systems store or record coordinates
in lat-long. There are various formats for presenting lat-long coordinates
(e.g. degrees minutes seconds, degrees decimal-minutes, decimal degrees). The most
consistent and easiest format to work with is *decimal degrees*. If you can collect data
in this format, then your analytical life will be much easier.

There are a nearly infinite number of possible projections. That said, there are a number of
standard projections that have been developed for various regions and scales. You should 
spend some time researching projections used in your region and adopt a single projection
that can serve as the standard for your analysis. In many cases the UTM zones may be a good
place to start. If you cannot find a projection that specifically meets your needs, you can
create a custom projection by adjusting parameters of an existing projection so the central
region is centered on your data.

So, why bother? Why not just keep your data in lat-long? For some applications (and, in the
spirit of keeping things simple), this may be a perfectly valid choice. However, lat-long
degrees do not represent equal distance and computation of spatial features and various
parameters are not as reliable. 

Thankfully, R and `sf` make it really easy to set and transform your data from one
projection to another. In our initial example, here, the projection information is
read directly from the shapefile.

```{r}
hiatus
```

Note there is both a `epsg (SRID)` entry and a `proj4string`. For standard projections,
the `epsg` code is the easiest and most reliable method for specifying a projection.
In some cases, however, you might need to specify a `proj4string`. If you don't know
the `epsg` or `proj4string` for your projection of interest, searching at 
(SpatialReference.org)[https://spatialreference.org] is a good place to start.

For South America, one possible projection to use is the South America Albers Equal 
Area Conic. The `epsg` integer code is 102033 and the PROJ.4 text description is:

```
+proj=aea +lat_1=-5 +lat_2=-42 +lat_0=-32 +lon_0=-60 +x_0=0 +y_0=0 +ellps=aust_SA +units=m no_defs
```

To transform our `hiatus` object into this projection we can simply execute the 
following

```{r}
hiatus <- hiatus %>% sf::st_transform(102033)
```


### Survey Stratum

In addition to the hiatus region, the survey area was divided into a northern and
southern stratum. The ShapeFiles for these polygons are provided in the same
directory as the hiatus data. The file names are: `Northern_stratum.shp` and
`Southern_stratum.shp`. Remember, you'll want to transform all of your
data into the South America Albers Equal Area Conic projection.

Before proceeding ahead, try creating a map on your own that includes all three
areas of interest (hiatus and both stratums).

```{r}
north_stratum <- file.path("..","data","Fraciscana_Example",
                           "Northern_stratum.shp") %>% 
  sf::st_read() %>% 
  sf::st_transform(102033)

# non-tidyverse
# north_path <- file.path("..","data","Fraciscana_Example",
#                         "Northern_stratum.shp")
# north_stratum <- sf::st_read(northern_path)

south_stratum <- file.path("..","data","Fraciscana_Example",
                           "Southern_stratum.shp") %>% 
  sf::st_read() %>% 
  sf::st_transform(102033)

# non-tidyverse
# north_path <- file.path("..","data","Fraciscana_Example",
#                         "Northern_stratum.shp")
# north_stratum <- sf::st_read(northern_path)

franciscana_plot <- 
  ggplot() +
  annotation_map_tile(type = esri_ocean, zoomin = 1, progress = "none") +
  layer_spatial(hiatus, col = "dark red", fill = "red", alpha = 0.4, size = 1) +
  layer_spatial(north_stratum, col = "purple4", fill = "purple", alpha = 0.4, size = 1) +
  layer_spatial(south_stratum, col = "purple4", fill = "purple", alpha = 0.4, size = 1) +
  geom_sf_label(north_stratum, mapping = aes(label = "Northern Stratum"), nudge_x = 150000) +
  geom_sf_label(south_stratum, mapping = aes(label = "Southern Stratum"), nudge_x = 165000) +
  geom_sf_label(hiatus, mapping = aes(label = "Hiatus"), nudge_x = 75000) +
  scale_x_continuous(expand = c(0.5,0.5)) +
  annotation_scale(location = "tl", 
                   width_hint = 0.3, 
                   style = "ticks") +
  ggtitle("Franciscana Survey Study Area",
          subtitle = "survey effort was concentrated in the northern and southern stratums")
franciscana_plot

```


## Load Spatial CSV Data into R

### Franciscana Sightings

```{r}
sightings <- file.path("..","data","Fraciscana_Example",
                           "Franciscana_Example_Sighting_Data.csv") %>% 
  readr::read_csv() 

sightings
```

As we look at this CSV file, we can see that each record represents a spatial point that
corresponds with a Franciscana sighting. Note that our coordinates are recorded in 
the `Lat` and `Long` columns and the units are decimal degrees. With this information we
can turn our `sightings` data into an sf object. Note that we will set our data initially
to `epsg` 4326. Then, we transform to our common projection.

```{r}
sightings <- sightings %>% 
  sf::st_as_sf(coords = c("Long","Lat")) %>% 
  sf::st_set_crs(4326) %>% 
  sf::st_transform(102033)
```


We will also be interested in distinguishing between "ON" and "OFF" effort
sightings. So, we will modify this column into a factor.

```{r}
sightings <- sightings %>% 
  dplyr::mutate(Effort = factor(Effort,levels = c("ON","OFF")))
```

```{r}
franciscana_plot + 
  layer_spatial(sightings) +
  ggtitle("Franciscana Survey Sightings", subtitle = NULL)
```

### Realized Survey Effort

In order to properly interpret the sightings data, we also need to know the survey effort
that was realized during the study. These data are provided as a CSV.

```{r}
realized_effort <- file.path("..","data","Fraciscana_Example",
                           "Franciscana_Example_Realized_Effort.csv") %>% 
  readr::read_csv()

realized_effort
```

If we examine these data, we see the structure provides four columns (`LatD_Beg`,
`LatD_End`, `LongD_Beg`, and `Long_End`) that represent our coordinates. These are
the start and end points for each line. Since these data actually represent lines,
we need to come up with a process for converting them to lines.

> These next few steps involve some intermediate to advance tidyverse skills. Don't worry
> if you don't grasp everything right away. We will take things slowly and step by
> step. The functions `gather` and `spread` within the `tidyr` package sometimes
> especially difficult for new users to fully appreciate.

The first step is we need to _tidy_ the data frame. One of the key principles of _tidy_
is that every record is an observation and ever column is a data field. In this 
situation, we have coordinates spread across 4 different columns. We would prefer to 
see a single column for `latitude` and another for `longitude`. Then, we would have
a third column that specifies the `label` (begin or end). From this, we could group the
points by Index and convert our `sf` geometry to "LINESTRING".


The first step is to use the `tidyr::gather()` function to make our data frame less 
wide. We are going to _gather_ the information in the four coordinate columns into 
two new columns `label`, and `coordinate`.

```{r}
realized_effort <- realized_effort %>% 
  gather(key = label, value = coordinate, LatD_Beg, LatD_End, LongD_Beg, LongD_End) %>% 
  tidyr::separate(label,c("longlat","beginend"),"_") %>% 
  dplyr::mutate(longlat = ifelse(longlat == "LatD","latitude","longitude"),
                beginend = ifelse(beginend == "Beg","begin","end")) %>% 
  dplyr::arrange(Index, LineLabel, Date)

realized_effort
```

Now, we can use the `tidyr::spread()` function to create a separate `longitude` and
`latitude` column to represent our coordinate data. We will then convert this into an
`sf` object. This will create a "POINT" data type, but we can use `dplyr::group_by()`
to identify our lines and then `summarise()` plus `sf::st_cast()` to create our final 
`sf` data table of effort lines.

```{r}
realized_effort <- realized_effort %>% 
  tidyr::spread(longlat,coordinate) %>% 
  sf::st_as_sf(coords = c("longitude","latitude")) %>% 
  sf::st_set_crs(4326) %>% 
  sf::st_transform(102033) %>% 
  dplyr::group_by(Index, LineLabel,Date) %>% summarise(do_union = FALSE) %>% 
  sf::st_cast("LINESTRING")

realized_effort
```

Lastly, lets add these lines to our previous map so we can see everything combined.

```{r}
franciscana_plot + 
  layer_spatial(realized_effort) +
  layer_spatial(sightings, mapping = aes(fill = Effort), size = 2, shape = 21) +
  scale_fill_manual(values = c("black","white"), drop = TRUE) +
  ggtitle("Franciscana Survey Sightings", 
          subtitle = "lines represent survey effort during the study")
```

## Multiple Panels in a Plot

In our previous example, it can be difficult to assertain any detail from the sightings
data at the larger scale. We might want to zoom in on each region and just show the
sightings within a particular region. This can be easily accomplished with the `sf` package.

```{r}
north_sightings <- sightings[north_stratum,]
hiatus_sigthings <- sightings[hiatus,]
south_sightings <- sightings[south_stratum,]

```

Now, we can create a plot that focuses on each region

```{r}
franciscana_north_plot <- 
  ggplot() +
  annotation_map_tile(type = esri_ocean, zoomin = 0, progress = "none") +
  layer_spatial(north_stratum, col = "dark red", fill = "red", alpha = 0.4, size = 1) +
  layer_spatial(realized_effort[north_stratum,]) +
  layer_spatial(sightings[north_stratum,], mapping = aes(fill = Effort), size = 3,
                shape = 21) +
  scale_fill_manual(values = c("black","white"), drop = FALSE) +
  scale_x_continuous(expand = c(0.25,0.25)) +
  annotation_scale(location = "tl", 
                   width_hint = 0.3, 
                   style = "ticks") +
  theme(legend.position="bottom") +
  ggtitle("Franciscana Sightings",
          subtitle = "northern stratum")
franciscana_north_plot
```

```{r}
franciscana_hiatus_plot <- 
  ggplot() +
  annotation_map_tile(type = esri_ocean, zoomin = 0, progress = "none") +
  layer_spatial(hiatus, col = "dark red", fill = "red", alpha = 0.4, size = 1) +
  layer_spatial(realized_effort[hiatus,]) +
  layer_spatial(sightings[hiatus,], 
                mapping = aes(fill = Effort),
                size = 3, shape = 21) +
  scale_fill_manual(values = c("black","white"), drop = FALSE) +
  scale_x_continuous(expand = c(0.25,0.25)) +
  annotation_scale(location = "tl", 
                   width_hint = 0.3, 
                   style = "ticks") +
  theme(legend.position="bottom") + 
  ggtitle("Franciscana Sightings",
          subtitle = "hiatus region")
franciscana_hiatus_plot
```

```{r}
franciscana_south_plot <- 
  ggplot() +
  annotation_map_tile(type = esri_ocean, zoomin = 0, progress = "none") +
  layer_spatial(south_stratum, col = "dark red", fill = "red", alpha = 0.4, size = 1) +
  layer_spatial(realized_effort[south_stratum,]) +
  layer_spatial(sightings[south_stratum,], 
                mapping = aes(fill = Effort), size = 3, shape = 21) +
  scale_fill_manual(values = c("black","white"), drop = FALSE) +
  scale_x_continuous(expand = c(0.25,0.25)) +
  annotation_scale(location = "tl", 
                   width_hint = 0.3, 
                   style = "ticks") +
  theme(legend.position="bottom") +
  ggtitle("Franciscana Sightings",
          subtitle = "southern stratum")
franciscana_south_plot
```

```{r}
library(gridExtra)

grid.arrange(franciscana_north_plot, franciscana_hiatus_plot,
             franciscana_south_plot,nrow = 1)
```

