--- 
title: "SOLAMAC Workshop - Spatial Analysis and Mapping in R"
author: "Josh M. London"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is a book of resources, references, and examples as part of the Spatial Analysis and Mapping in R workshop as part of SOLAMAC 2018 in Lima, Peru"
---

# Prerequisites

For best results, the user should have the latest version of R (https://www.r-project.org/)
and RStudio (https://www.rstudio.com/products/rstudio/download/) installed.

The next step is to install a variet of packages that will be needed for spatial
analysis and creating maps in R. We will start with those packages available
from CRAN.

```{r install-pkgs, eval = FALSE}
install.packages(c(
  "tidyverse",
  "sf",
  "devtools"
))
```

The package `rgeos` currently has a bug that prevents proper installation on some
systems. So, we will use the `rgeos::install_version()` function to install
the previous version.

```{r install-rgeos, eval = FALSE}
devtools::install_version("rgeos", version = "0.3-28")
```

Lastly, we will install the `ptolemy` package which provides access to the
Global High-Resolution Geography for shoreline data. This will provide easy
access to basemaps of global land masses. The package is not released to CRAN
and so we will use `devtools::install_github()`

```{r install-ptolemy, eval = FALSE}
devtools::install_github('jmlondon/ptolemy')
# if asked, do not upgrade the rgeos install
```

The `ptolemy` package needs to download and install the source data before
you can access the data.

```{ install-gshhg, eval = FALSE}
ptolemy::install_gshhg()
```

At this point, you should be ready to proceed with the examples provided in this
book. Each chapter is a separate example and each tries to introduce you to 
elements of the `sf` and `ggplot2` / `ggspatial` packages that make importing,
manipulating, and creating maps in R relatively easy.
